SMT-Boosted Security Types for Low-Level MPC

### 2025-11-01 toplas26/metalang.tex
checkpoint

The $\metaprot$ language \cite{skalka-near-ppdp24} includes structured
data and function definitions for defining composable protocol
elements at a higher level of abstraction than $\minicat$.  The
$\metaprot$ language is a \emph{metalanguage} aka metaprogramming
language, where an $\fedprot$ protocol is the result of
computation. In addition to these declarative benefits of $\metaprot$,
component definitions support compositional verification of larger
protocols. Separate verification of well-designed components results
in confidentiality and integrity properties to be recorded in their
types, allowing for significant reduction of SMT verification in whole
program analysis, as we will discuss with extended examples in Section
\ref{section-examples}.

@@

The syntax of $\metaprot$ is defined in Figure
\ref{fig-metaprotsyntax}.  It includes a syntax of values $\mv$ that
include client ids $\cid$, identifier strings $w$, expressions $\be$
in field $\mathbb{F}_p$, record values, and $\minicat$ variables
$x$. $\metaprot$ expression forms allow dynamic construction of these
values. $\metaprot$ \emph{instruction} forms allow dynamic
construction of $\minicat$ protocols $\prog$ that incorporate expression
evaluation. The syntax also supports definitions of functions that
compute values $\mv$ and, as a distinct form, functions that compute
protocols $\prog$.  Formally, we consider a complete metaprogram to include
both a codebase and a ``main'' program that uses the codebase.
We disallow recursion, mainly to guarantee decidability
of type dependence (Section \ref{section-metalangty}).
\begin{definition}
A \emph{codebase} $\codebase$ is a list of function 
declarations. We write $ \codebase(f) = y_1,\ldots,y_n,\ e$
if $f(y_1,\ldots,y_n) \{ e \} \in \codebase$, and we
write  $ \codebase(f) = y_1 : \tau_1, \ldots, y_n : \tau_n, \instr$
if $f(y_1 : \tau_1, \ldots, y_n : \tau_n) \{ \instr \} \in \codebase$.
%A \emph{metaprogram}, aka \emph{metaprotocol}  is a pair of a 
%codebase and expression $\codebase, e$. We may omit
%$\codebase$ if it is clear from context.  
\end{definition}

@@

We define a big-step evaluation relation $\redx$ in Figures
\ref{fig-metaprotexprsemantics} and \ref{fig-metaprotinstrsemantics}
for expressions and instructions, respectively.  In this definition we
write $e[\mv/y]$ and $\instr[\mv/y]$ to denote the substitution of
$\mv$ for free occurrences of $y$ in $e$ or $\instr$ respectively. The
rules are mostly standard. Note that we do not include an evaluation
rule for the form $\eqcast{\mx{e}{e}}{\notg{\phi}}$ which is a type
annotation that we assume is erased from programs prior to
evaluation. We defer discussion of this form, as well as the syntactic
category $\notg{\phi}$, to the next Section.

@@

\subsection{$\metaprot$ Automated Abstract Hoare Logic}

\begin{verbatimtab}
precondition: ( m[x++"foo"]@1 == 3 )
bar(x:string){ m[x++"bar"]@2 := (m[x++"foo"] + m[x++"baz"])@1 }
postcondition: ( [m[z++"bar"]@2 == m[x++"baz"]@1 + 3 )
\end{verbatimtab}

@@

@@

\begin{fpfig}[t]{Algorithmic $\metaprot$ Abstract Hoare Triple Deduction Rules}{fig-metahtrip}
\begin{mathpar}
\inferrule[GenEntails]
          {\mv_1,\ldots,\mv_n = \fresh(\tau_1,\ldots,\tau_n) \\ \peq_1[\mv_1/y_n \cdots \mv_n/y_n] \redx \eqs_1 \\
           \peq_2[\mv_1/y_n \cdots \mv_n/y_n] \redx \eqs_2 \\ \eqs_1 \models \eqs_2}
          {\forall y_1:\tau_1,\ldots,y_n:\tau_n . \peq_1 \models \peq_2 }

  \inferrule[Mesg]
            {}
            {\htrip{\eqtrue}{\xassign{e_1}{e_2}{e_3}}{e_1 \eop \toeq{\elab{e_2}{e_3}}}}

  %\inferrule[Encode]
  %          {\mx{e_1}{e_2} \redx x \\ \notg{\phi} \redx \phi \\
  %            \eqs \models x \eop \phi\\
  %            \atj{\phi}{R}{\ty}}
  %          {\mtj{\eqcast{\mx{e_1}{e_2}}{\notg{\phi}}}{\eqs}{(x : \ty)}{R}{\varnothing}{\eqs}}
  %
  \inferrule[Assert]
            {}
            {\htrip{\toeq{\elab{e_1}{e_3}} \eop \toeq{\elab{e_2}{e_3}}}{\elab{\assert{e_1 = e_2}}{e_3}}{\eqtrue}}

  \inferrule[Seq]          
            {\htrip{\peq_1^1}{\cmd_1}{\peq_2^1} \\ \htrip{\peq_1^2}{\cmd_2}{\peq_2^2}}
            {\htrip{\peq_1^1 \wedge (\peq_2^1 \impl \peq_1^2)}{\cmd_1;\cmd_2}{\peq_2^1 \wedge \peq_2^2}}

  \inferrule[Let]
            {\htrip{\peq_1}{\cmd[e/y]}{\peq_2}}
            {\htrip{\peq_1}{\elet{y}{e}{\cmd}}{\peq_2}}

  \inferrule[App]
            {\htrip{\peq_1}{f(y_1:\tau_1,\ldots,y_n:\tau_n)}{\peq_2}}
            {\htrip{\peq_1[e_1/y_1 \cdots e_n/y_n]}{f(e_1,\ldots,e_n)}{\peq_2[e_1/y_1 \cdots e_n/y_n]}}

  \inferrule[Fn]
            {\codebase(f) = y_1 : \tau_1, \ldots, y_n : \tau_n, \cmd \\ \htrip{\peq_1}{\cmd}{\peq_2}}
            {\htrip{\peq_1}{f(y_1:\tau_1,\ldots,y_n:\tau_n)}{\peq_2}}

  \inferrule[Hardpack]
            {\precond(f) = \peq_1 \\ \postcond(f) = \peq_2 \\
              \htrip{\peq_1'}{f(y_1:\tau_1,\ldots,y_n:\tau_n)}{\peq_2'} \\
              \forall y_1:\tau_1,\ldots,y_n:\tau_n . 
            \peq_1 \models \peq_1'\ \text{and}\ \peq_1 \wedge \peq_2' \models \peq_2}
            {\htrip{\peq_1}{f(y_1:\tau_1,\ldots,y_n:\tau_n)}{\peq_2}}
\end{mathpar}
\end{fpfig}

\begin{lemma}
  \label{lemma-closure}
  If $\htrip{\peq_1}{\prog}{\peq_2}$ and $\instr[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \prog$
  then $\peq_1[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \eqs_1$ and
  $\peq_2[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \eqs_2$ for some $\eqs_1, \eqs_2$. 
\end{lemma}

\begin{lemma}[$\metaprot$ Hoare Triple Correctness]
  \label{lemma-htrip}
  If $\htrip{\peq_1}{\cmd}{\peq_2}$ with $\peq_1 \redx
  \eqs_1$, $\peq_2 \redx \eqs_2$, and $\cmd \redx
  \prog$, then $\htrip{\eqs_1}{\prog}{\eqs_2}$.
\end{lemma}

@@

    \htrip{\eqs_1^2}{\cmd_2}{\eqs_2^2}
  \end{mathpar}
  But by \TirName{Hyp} we have:
  $\htrip{\eqs_1^1 \wedge (\eqs_2^1 \impl \eqs_1^2)}{\cmd_1}{\eqs_2^1 \wedge \eqs_1^2}$
  and by sanity conditions \cnote{need to clarify these} we have
  $\vars(\eqs_2^1) \cap \avars(\prog_2) = \varnothing$ so by \TirName{Frame} we
  have
  $\htrip{\eqs_2^1 \wedge \eqs_1^2}{\cmd_2}{\eqs_2^1 \wedge \eqs_2^2}$.
  Thus by $\TirName{Seq}$ we have:
  $$\htrip{\eqs_1^1 \wedge (\eqs_2^1 \impl \eqs_1^2)}{\prog_1;\prog_2}{\eqs_2^1 \wedge \eqs_2^2}$$

@@

\begin{theorem}
  If $\htrip{\eqs_1}{\ttt{main()}}{\eqs_2}$ as a consequence of
  \TirName{HardPack} and $\ttt{main}() \redx \prog$ then
  $\htrip{\eqs_1}{\prog}{\eqs_2}$.
\end{theorem}

>>>>

The $\metaprot$ language \cite{skalka-near-ppdp24} includes structured
data and function definitions for defining composable protocol
elements at a higher level of abstraction than $\minicat$.  The
$\metaprot$ language is a \emph{metalanguage} aka metaprogramming
language, where an $\fedprot$ protocol is the result of
computation. In addition to these declarative benefits of $\metaprot$,
component definitions support compositional verification of larger
protocols. This includes an automated verification of algebraic
program properties expressed as Hoare logic assertions that we
introduce in this Section. Furthermore, we can compositionally verify
confidentiality and integrity properties expressed as types, as we
will show in Section \ref{section-sharety}. We present and discuss
extended examples of $\metaprot$ programs in Section
\ref{section-examples}.

@@

The syntax of $\metaprot$ is defined in Figure
\ref{fig-metaprotsyntax}.  It includes a syntax of values $\mv$ that
include client ids $\cid$, identifier strings $w$, expressions $\be$
in field $\mathbb{F}_p$, record values, and $\minicat$ variables
$x$. $\metaprot$ expression forms allow dynamic construction of these
values. Note in particular that $\concat$ is string concatentation.
$\metaprot$ \emph{instructions} allow dynamic
construction of $\minicat$ protocols $\prog$ that incorporate expression
evaluation. The syntax also supports definitions of functions that
compute values $\mv$ and, as a distinct form, functions that compute
protocols $\prog$.  The latter requires type annotations on arguments
which are restricted to being either strings, client ids, or
records-- we use this type information to support Hoare logic
automation as we discuss below in Section \ref{section-metahtrip}.

Formally, we consider a complete metaprogram to include
both a codebase and a ``main'' program that uses the codebase.
We disallow recursion, mainly to guarantee decidability
of type dependence (Section \ref{section-metalangty}).
\begin{definition}
A \emph{codebase} $\codebase$ is a list of function 
declarations. We write $ \codebase(f) = y_1,\ldots,y_n,\ e$
if $f(y_1,\ldots,y_n) \{ e \} \in \codebase$, and we
write  $ \codebase(f) = y_1 : \tau_1, \ldots, y_n : \tau_n, \instr$
if $f(y_1 : \tau_1, \ldots, y_n : \tau_n) \{ \instr \} \in \codebase$.
%A \emph{metaprogram}, aka \emph{metaprotocol}  is a pair of a 
%codebase and expression $\codebase, e$. We may omit
%$\codebase$ if it is clear from context.  
\end{definition}

@@

We define a big-step evaluation relation $\redx$ in Figures
\ref{fig-metaprotexprsemantics} and \ref{fig-metaprotinstrsemantics}
for expressions and instructions, respectively.  In this definition we
write $e[\mv/y]$ and $\instr[\mv/y]$ to denote the substitution of
$\mv$ for free occurrences of $y$ in $e$ or $\instr$ respectively. The
rules are mostly standard.

Note that we do not include an evaluation
rule for the form $\eqcast{\mx{e}{e}}{e}$ which is a type
annotation that we assume is erased from programs prior to
evaluation. We defer discussion of this form to Section
\ref{section-sharety}

\paragraph{Sanity conditions on programs.} We assume that
any $\metaprot$ codebase $\codebase$ includes a function
$\ttt{main}()$. And as for $\minicat$, we assume that $\metaprot$
programs are semantically well-formed-- that is, that
$\ttt{main}() \redx \prog$ and $\prog$ is well-formed.

@@

\subsection{Abstract pre and postcondition annotations}

$\metaprot$ instruction functions can optionally be annotated with
pre and postconditions. To model this in our theory we posit
functions $\precond$  and $\postcond$ 
that specify conditions for instruction functions $f \in \dom(\codebase)$.
Note that $\precond(f)$ or $\precond(f)$ may be undefined if no annotation
is given for $f$. In actual code, we write pre and postconditions directly before and
after functions to define them. Here is a simple example function $\ttt{bar}$
with. Note that the pre and postconditions are abstract with respect to the
input string $\ttt{x}$, and that string concatenation  
can be used in pre and postcondition constraints:
\begin{verbatimtab}
precondition: ( m[x++"foo"]@1 == 3 )
bar(x:string){ m[x++"bar"]@2 := (m[x++"foo"] + m[x++"baz"])@1 }
postcondition: ( [m[z++"bar"]@2 == m[x++"baz"]@1 + 3 )
\end{verbatimtab}
Since expressions can occur in $\metaprot$ constraints, they are not
SMT constraints, but abstractions of constraints parameterized by
the formal parameters of the function they annotate. We denote these
\emph{abstract constraints} as $\eqs$ with the syntax given
in Figure \ref{fig-peq}. When the function is applied, the pre and postconditions
are instantiated as SMT constraints, via evaluating the expressions
within the constraints. The evaluation semantics of abstract constraints
are also given in Figure \ref{fig-peq}.

@@

In the case of \ttt{main}, since it has no arguments, the
pre and postconditions must be concrete SMT expressions.
For example, given the \ttt{bar} function above, we
could define a full program as follows, where \ttt{T}
is the true constraint. 
\begin{verbatimtab}
precondition: ( T )
main(){
  m["thisfoo"]@1 := 3; m["thisbaz"]@1 := 2; bar("this")
}
postcondition: ( [m["thisbar"]@2 == 5 )
\end{verbatimtab}
There are no preconditions to this \ttt{main} because they are
all established by the program, and the postcondition follows
as a consequence of the application \ttt{bar("this")}.

The use of pre and postconditions in \ttt{main} allows verification of
whole-program (protocol) correctness. Further, the use of annotations
on function definitions supports \emph{compositional} verification of
program components. In general postconditions can serve as
pre-verified properties of functions that can be reused in, and reduce
overhead of, whole protocol verification.

\subsection{$\metaprot$ Automated Abstract Hoare Logic}
\label{section-metahtrip}

In the $\metaprot$ theory, we introduce Hoare logic assertions
of the form $\htrip{\peq_1}{\cmd}{\peq_2}$. The correctness
condition of this assertion is compositional, in that instantiated
abstract Hoare triples evaluate to valid $\minicat$ Hoare triples.
This is formally elaborated below in Lemma
\ref{lemma-metahtrip} and Corollaries \ref{cor-metahtrip} and
\ref{cor-mainhtrip} below.

In Figure \ref{fig-metahtrip} we give the derivation rules for the
abstract Hoare logic. These are based on $\minicat$ Hoare logic, but
are algorithmic. A main predicate for these rules is the
$\TirName{GenEntails}$ rule, which encodes a generalized version of
entailment. By using fresh variables to instantiate abstract
constraints, we guarantee entailment for any instantiation, as
follows.

@@

\metahtripfig

The \TirName{HardPack} rule uses \TirName{GenEntails} to ensure that
the annotated pre and postconditions are correct for a given function,
if they are defined. It is in fact not necessary for \emph{any} pre
and postconditions to be defined, but their intent is to enable
fine-tuning of postconditions for improving verification efficiency as
discussed above and later in Section \ref{section-examples}.  While
not relevant to the theory, an important implementation detail is that
verification of any given function $f$ only needs to be performed
once, and can then be memoized for other applications using
the $\TirName{App}$ rule as a consequence of Lemma \ref{lemma-genentails}
(and more precisely of Corollary \ref{cor-metahtrip}).

The \TirName{Assign} and \TirName{Assert} rules propagate equations
induced by assignments and assertions to post and preconditions
respectively.  Note that without the use of \TirName{HardPack}, the
entire protocol would accumulate in postconditions of \ttt{main}. The
\TirName{Seq} rule relies on the correctness of our \TirName{Hyp} rule
to join pre and postconditions as we discuss in the relevant case of
Lemma \ref{lemma-metahtrip}, which is the foundational correctness
result for the analysis.

\paragraph{Pre and post sanity conditions.} We require that
annotated pre and postconditions on functions will evaluate to SMT
constraints when instantiated with function arguments, and refer only
to variables that occur in their bodies. Deduction rules other than
\TirName{HardPack} enforce this by definition.  For any $f$ with
$\codebase(f) = y_1 : \tau_1, \ldots, y_n : \tau_n, \cmd$,
$\precond(f) = \peq_1$, and $\postcond(f) = \peq_2$, we require that
for all $[\mv_1/y_1]\cdots[\mv_n/y_n]$ there exists $\eqs_1$, $\eqs_2$
where:
\begin{mathpar}
  \cmd[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \prog

  \peq_1[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \eqs_1

  \peq_2[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \eqs_2
\end{mathpar}
and $\vars(\eqs_1) \cup \vars(\eqs_2) \subseteq \vars(\prog)$.

\subsection{Abstact Hoare Logic Soundnes}

We begin by showing that the analysis preserves the property that in
any $\htrip{\peq_1}{\prog}{\peq_2}$, fully instantiating $\prog$
also fully instantiates $\peq_1$ and $\peq_2$. The result
follows by sanity conditions on pre and postcondition annotations,
the definition of analysis, and induction on $\cmd$.
\begin{lemma}
  \label{lemma-closure}
  If $\htrip{\peq_1}{\prog}{\peq_2}$ and $\instr[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \prog$
  then $\peq_1[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \eqs_1$ and
  $\peq_2[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \eqs_2$ for some $\eqs_1, \eqs_2$. 
\end{lemma}

Now we can prove our main result in a general form. The proof
follows mainly by appeal to the $\minicat$ Hoare logic deduction
system.
\begin{lemma}[$\metaprot$ Hoare Triple Correctness]
  \label{lemma-metahtrip}
  If $\htrip{\peq_1}{\cmd}{\peq_2}$ with $\peq_1 \redx
  \eqs_1$, $\peq_2 \redx \eqs_2$, and $\cmd \redx
  \prog$, then $\htrip{\eqs_1}{\prog}{\eqs_2}$.
\end{lemma}

@@

    \htrip{\eqs_1^2}{\cmd_2}{\eqs_2^2}
  \end{mathpar}
  But by \TirName{Hyp} we have:
  $\htrip{\eqs_1^1 \wedge (\eqs_2^1 \impl \eqs_1^2)}{\cmd_1}{\eqs_2^1 \wedge \eqs_1^2}$
  and by sanity conditions we have
  $\vars(\eqs_2^1) \cap \avars(\prog_2) = \varnothing$ so by \TirName{Frame} we
  have
  $\htrip{\eqs_2^1 \wedge \eqs_1^2}{\cmd_2}{\eqs_2^1 \wedge \eqs_2^2}$.
  Thus by $\TirName{Seq}$ we have:
  $$\htrip{\eqs_1^1 \wedge (\eqs_2^1 \impl \eqs_1^2)}{\prog_1;\prog_2}{\eqs_2^1 \wedge \eqs_2^2}$$

@@

As a consequence of the preceding Lemma, we can state important
Corollaries that characterize, first, the compositionality of function
constraints, and second, the correctness of whole program
analysis. These both follow immediately by reasoning in the
\TirName{App} case of Lemma \ref{lemma-metahtrip}.

\begin{corollary}[Compositionality of Abstract Function Constraints]
  \label{cor-metahtrip}
  Given:
  \begin{mathpar}
    \htrip{\peq_1}{f(y_1:\tau_1,\ldots,y_n:\tau_n)}{\peq_2}

    \codebase(f) = y_1 : \tau_1, \ldots, y_n : \tau_n, \cmd

    \cmd[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \prog

    \peq_1[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \eqs_1

    \peq_2[\mv_1/y_1]\cdots[\mv_n/y_n] \redx \eqs_2
  \end{mathpar}
  Then $\htrip{\eqs_1}{\prog}{\eqs_2}$ is valid.
\end{corollary}

\begin{corollary}
  \label{cor-mainhtrip}
  If $\htrip{\peq_1}{\ttt{main()}}{\peq_2}$ and $\ttt{main}() \redx \prog$ and
  $\peq_1 \redx \eqs_1$ and $\peq_2 \redx \eqs_2$ then
  $\htrip{\eqs_1}{\prog}{\eqs_2}$.
\end{corollary}
